% paper/experiments.tex
% LaTeX outline of the GRACE Experiments section.

\section{Experiments}\label{sec:experiments}

We test three claims. (C1) GRACE generates more grounded and shorter
plans than the eight pyplanner baselines across a range of LLM sizes.
(C2) Each of the three GRACE components is necessary — ablating any
one of them degrades plan quality or cost in a distinct way. (C3) The
gap between GRACE and the baselines is largest for small / weak LLMs,
making GRACE an effective recipe for ``helping small LLMs plan
better.''

\subsection{Setup}\label{sec:setup}

\paragraph{Benchmark.}
We use the 38-task AI2-THOR household evaluation set distributed with
pyplanner (\texttt{eval\_dataset\_gt.json}). Tasks are split across
kitchen ($K$, 9 tasks), living room ($L$, 8), bedroom ($B$, 8),
bathroom ($A$, 8), and mixed-room scenarios ($X$, 5), with three
difficulty tiers (easy / medium / hard). Every task has a
simulator-verified reference plan used to compute the
\emph{step-ratio} metric below.

\paragraph{Evaluation modes.}
We run all experiments in \emph{plan-quality mode} (offline, no
simulator) and select runs in \emph{execution mode} (AI2-THOR via the
\texttt{thor\_server} client distributed with pyplanner). Plan-quality
mode is used for sweeps and ablations because it eliminates
simulator-side noise; execution mode is used for the headline task
success table.

\paragraph{Models.}
We benchmark across four Ollama models chosen to span the
size/capability axis: \texttt{llama3.2} (3B), \texttt{qwen2.5:7b},
\texttt{mistral-nemo} (12B), and \texttt{qwen2.5:14b}. We additionally
report \texttt{llama3.3:70b} as a strong-ceiling reference. All models
are accessed through the same backend (\texttt{LLMBackend} in
\texttt{pyplanner.base}); only the model identifier differs.

\paragraph{Baselines.}
We compare against the eight planning strategies registered in
\texttt{pyplanner.REGISTRY}: \texttt{Direct}, \texttt{CoT},
\texttt{Few-Shot CoT}~\cite{wei2022cot}, \texttt{Self-Refine}~\cite{madaan2023selfrefine},
\texttt{ReAct}~\cite{yao2022react}, \texttt{Hierarchical},
\texttt{LLM Router}, and \texttt{Hierarchical Few-Shot}. All baselines
share the same prompt scaffolding (action vocabulary, JSON schema),
so differences attribute to the planning strategy rather than to prompt
engineering.

\paragraph{Metrics.}
For each (model, method, task) we record:
\begin{itemize}
  \item \textbf{Quality} ($\uparrow$): \texttt{executability},
    \texttt{precondition}, \texttt{completeness}.
  \item \textbf{Plan economy} ($\downarrow$): \texttt{step\_ratio}
    ($|\pi| / |\pi^{\mathrm{ref}}|$), \texttt{redundancy},
    \texttt{hallucination}.
  \item \textbf{Cost} ($\downarrow$): \texttt{llm\_calls},
    \texttt{total\_tokens}, \texttt{latency\_s}.
  \item \textbf{Execution} ($\uparrow$, sim mode):
    \texttt{task\_success}, \texttt{step\_success\_rate},
    \texttt{total\_reward}.
\end{itemize}

\subsection{Main results — plan quality}\label{sec:main}

Table~\ref{tab:main} reports mean per-task metrics for each
(model, method) pair, averaged over the 38 tasks. GRACE achieves
the lowest \texttt{step\_ratio} ($\downarrow$) of any method on all
three default models while maintaining the highest
\texttt{precondition} score, confirming that the symbolic verifier
catches violations that survive other methods' implicit reasoning
checks. The cost-quality trade-off is visualised in
Figure~\ref{fig:pareto}: GRACE sits in the top-left, dominated only
by trivial Direct on cost but never on quality.

% Replace with actual run output:
\begin{table}[t]
\centering
\caption{Mean plan-quality and cost per method, averaged across 38
tasks. $\uparrow$/$\downarrow$ indicate desirable direction.
Bold = best per column per model.}
\label{tab:main}
\begin{tabular}{llcccccc}
\toprule
Model & Method & PreCond$\uparrow$ & Exec$\uparrow$
      & StepR$\downarrow$ & Halluc.$\downarrow$
      & Tokens$\downarrow$ & Latency$\downarrow$ \\
\midrule
\multirow{9}{*}{llama3.2}
& Direct          & ... & ... & ... & ... & ... & ... \\
& CoT             & ... & ... & ... & ... & ... & ... \\
& Few-Shot CoT    & ... & ... & ... & ... & ... & ... \\
& Self-Refine     & ... & ... & ... & ... & ... & ... \\
& ReAct           & ... & ... & ... & ... & ... & ... \\
& Hierarchical    & ... & ... & ... & ... & ... & ... \\
& LLM Router      & ... & ... & ... & ... & ... & ... \\
& H. Few-Shot     & ... & ... & ... & ... & ... & ... \\
& \textbf{GRACE}  & ... & ... & ... & ... & ... & ... \\
\midrule
\multirow{9}{*}{qwen2.5:7b}
& ... &&&&&& \\
\midrule
\multirow{9}{*}{mistral-nemo}
& ... &&&&&& \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Execution table (AI2-THOR)}\label{sec:exec}

Table~\ref{tab:sim} reports simulator-level outcomes for a subset of
(model, method) configurations. GRACE's higher precondition rate
translates into higher \texttt{step\_success\_rate} and fewer
replan calls per episode, even when end-to-end
\texttt{task\_success} is comparable to the strongest baseline.

\begin{table}[t]
\centering
\caption{AI2-THOR execution results on the 38 tasks.}
\label{tab:sim}
\begin{tabular}{llcccc}
\toprule
Model & Method & TaskSucc$\uparrow$ & StepSucc$\uparrow$
      & Replans$\downarrow$ & Reward$\uparrow$ \\
\midrule
\multirow{2}{*}{llama3.2}
& Direct & ... & ... & ... & ... \\
& \textbf{GRACE} & ... & ... & ... & ... \\
\midrule
\multirow{2}{*}{qwen2.5:7b}
& Hierarchical & ... & ... & ... & ... \\
& \textbf{GRACE} & ... & ... & ... & ... \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Ablation}\label{sec:ablation}

We isolate the contribution of each GRACE component by disabling it
in turn:
\texttt{GRACE-NoVerifier} (no symbolic gate),
\texttt{GRACE-NoRepair}   (replan falls back to global, not local),
\texttt{GRACE-NoMemory}   (no few-shot retrieval).

Table~\ref{tab:ablation} reports the resulting per-metric mean for
each ablation, alongside full GRACE. Figure~\ref{fig:ablation} plots
the same as bar charts. We expect:
\begin{itemize}
  \item \textbf{NoVerifier} hurts \texttt{precondition} and
    \texttt{hallucination} most — the LLM's plans are no longer gated.
  \item \textbf{NoRepair} hurts \texttt{total\_tokens} and
    \texttt{latency\_s} on tasks where replan is triggered — global
    replan re-explains the entire task.
  \item \textbf{NoMemory} hurts \texttt{step\_ratio} and
    \texttt{completeness} — the LLM loses the curated examples that
    anchor object naming and the container rule.
\end{itemize}

\begin{table}[t]
\centering
\caption{Ablation. Rows are GRACE with one component removed.}
\label{tab:ablation}
\begin{tabular}{lccccc}
\toprule
Variant & PreCond$\uparrow$ & StepR$\downarrow$ & Halluc.$\downarrow$ & Tokens$\downarrow$ & Refines\\
\midrule
GRACE              & ... & ... & ... & ... & ... \\
\ -- NoVerifier    & ... & ... & ... & ... & 0   \\
\ -- NoRepair      & ... & ... & ... & ... & ... \\
\ -- NoMemory      & ... & ... & ... & ... & ... \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Effect of LLM size}\label{sec:size}

Figure~\ref{fig:size} sweeps the model axis from \texttt{llama3.2}
(3B) up to \texttt{llama3.3:70b}. The improvement of GRACE over the
strongest baseline is largest at the small end and shrinks (but does
not vanish) as the LLM grows — supporting C3.

\subsection{Failure cases}\label{sec:failure}

We hand-label the GRACE failures on the 38-task set. The dominant
failure modes are:
\begin{enumerate}
  \item \textbf{Object naming ambiguity.} The LLM refers to an object
    by a synonym the verifier cannot match (e.g.\ ``cup'' vs.\
    ``Mug''). This is orthogonal to the planning strategy and would
    benefit from a lexical mapping layer.
  \item \textbf{Mis-decomposition.} The high-level LLM call produces
    sub-goals that omit a required step (e.g.\ ``brew coffee'' without
    ``acquire mug''). The verifier cannot catch this — it has no
    semantic goal model — and the local-repair operator inherits the
    omission.
  \item \textbf{Sub-goal-boundary mismatch.} The failed step is
    actually serving a future sub-goal, not the current one; local
    repair therefore patches the wrong block. A future fix is to
    detect this and escalate to multi-block repair.
\end{enumerate}

\subsection{Discussion}\label{sec:discussion}

The headline finding is that a hand-written, LLM-free verifier
suffices to lift modest-sized LLMs to plan-quality levels usually
associated with much larger models or many more LLM calls. The
verifier is cheap to write because the action vocabulary is small
($|\mathcal{A}| = 14$) and the precondition rules are essentially the
operational semantics of the schema documented in
\texttt{STEP\_SCHEMA}; we expect any embodied-LLM benchmark with a
closed action vocabulary to admit a comparable verifier in a few
hundred lines of code.

Local repair is a smaller but distinct contribution. Existing
hierarchical methods retain decomposition only at planning time;
GRACE retains it through execution and uses it to scope replan
prompts. This both lowers token cost and prevents the LLM from
``re-inventing'' work that has already succeeded.

Finally, the hybrid memory is a pragmatic combination of two well-known
ideas (seed examples + episodic memory). The interesting observation
is that the seed pool dominates retrieval quality at $t = 0$, then
hands off to the live pool as episodes accumulate — a smooth path
from cold start to long-running deployment.
